Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity\vspace{-7ex}

### 2016-08-12 papers/kmeans/kmeans.tex
ready to resubmit

\begin{abstract}
% Context:
Chemical tagging promises to use detailed abundance measurements to
identify spatially separated stars that were in fact born together
(in the same molecular cloud), long ago.
This idea has not yielded much success, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic
measurements with \thecannon, which has now delivered $15$ individual
abundances for $\sim 10^5$ stars observed as part of the
\apogee\ spectroscopic survey, with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars
in the 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known phase-space clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
This is the first-ever project to identify phase-space structures at survey-scale by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

In what follows, we are going to use \apogee\ \acronym{DR12}
data \citep{apogeedr12, dr12}
from \sdssiii\ \citep{sdssiii}, in
which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---is capable of exploring all parts of the disk,
including the thin parts.
However, it has the disadvantage that its spatial coverage is incomplete
(that is, limited pointings produce a fractured sky map),
which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage; that
said, it will not observe much of the thin disk.

@@

Another issue with k-means is that it effectively uses metric
distances in the $D$-space; it presumes Euclidean isotropy.
We choose here to work in the hydrogen-normalized abundance space, the
space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
But in addition to this, we re-scale these by approximate typical measurement
precisions obtained by \thecannon\ before running k-means.
These scalings were
  [C/H]/0.041,
  [N/H]/0.044,
  [O/H]/0.037,
  [Na/H]/0.111,
  [Mg/H]/0.032,
  [Al/H]/0.055,
  [Si/H]/0.041,
  [S/H]/0.054,
  [K/H]/0.069,
  [Ca/H]/0.043,
  [Ti/H]/0.072,
  [V/H]/0.146,
  [Mn/H]/0.041,
  [Fe/H]/0.019, and
  [Ni/H]/0.034.
This scaling makes the space close to isotropic in measurement uncertainty or
observational precision.
Importantly we use \emph{only} abundance-space information, and no
positional or velocity information (nor $\teff$ nor $\logg$ nor any
targeting or observational meta-data) as input to the clustering
algorithm.

@@

We find a mean metallicity of [Fe/H]$ = -1.3$ for the cluster we associate as M~5 (\figurename~\ref{fig:M5}), which agrees well with the [Fe/H]$ = -1.33 \pm 0.03$ measurement reported by \citet{Koch_2010}, and the $-1.29 \pm 0.02$ value listed in the compilation of \citet[][accessed 2016]{Harris_1996}.  The correlations in light elements---specifically Mg--Al and C--N---also agree well with other studies \citep{Ivans_2001,meszaros}.  In particular we find only a weak correlation in the [Na/Fe]--[O/Fe] abundance ratios \citep{Lai_2011}.  However, when we consider the extent of the literature on M~5, it would suggest that we do not recover the full extent of these correlations: the stars with the highest [Na/Fe], [Mg/Al], and lowest [O/Fe] abundance ratios are not represented in the cluster that we associate as M~5.  This is likely a consequence of our (poor) choice of the k-means algorithm, which is most effective for near-normal distributions, and clearly less powerful for clusters (in a statistical sense) that have polynomial relationships in dimensional space (e.g., [Na/Fe]--[O/Fe]).  Indeed, \citet{meszaros} report abundances for 122 members of M~5 using \apogee\ \acronym{DR12} data, whereas the cluster we associate as M~5 only contains $\approx60$ members, and only covers about half of the extent of [Na/Fe]--[O/Fe] relationship.

@@

It is conceivable that this structure, being near the outskirts of the
abundance distribution, is somehow concentrated in abundance space by
some kind of shrinkage (in the statistical sense) induced by
\thecannon's regression.
However, it is a very dense feature in abundance space, and is worth
following up.
If follow-up observations show that the stars have abundances and that
are consistent with having origin in a single stellar population or
low-metallicty dwarf galaxy, it could represent an accretion event in
the Milky Way halo.

The k-means algorithm employed here is by no means optimal, and other clusters
we identify are accompanied by a few stars that are not currently co-spatial.
Those stars may be unassociated interlopers that have been misclassified,
or they may be true cluster members that are now unbound. However in the
case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
the situation is far more extreme. We identify a group of stars with
very similar abundances in 15 dimensions that are now spread throughout
the Galaxy halo.
Because some stars ($<8\%$) are candidate members of the massive globular
cluster M53, it provides the tantalizing possibility that these stars may
indeed be all members of the same co-natal gas cloud, implying that 
these stars have been accreted onto the halo from a single globular
cluster early in the Milky Way's formation. While speculative,
this idea demonstrates the promise of chemical tagging.

@@

\acknowledgements
It is a pleasure to thank an anonymous referee for useful comments that have led to improvements in the manuscript.
We also thank
  Joss Bland-Hawthorn (Sydney),
  Jo Bovy (Toronto),
  Charlie Conroy (Harvard),
  Katia Cunha (\acronym{NOAO}),
  Amina Helmi (Kapteyn),
  Jeremy Magland (\acronym{SCDA}),
  Don Schneider (\acronym{PSU}),
  Keivan Stassun (Vanderbilt),
  Angus Williams (Cambridge),
  and the Blanton--Hogg group meeting
for valuable discussions and comments.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1312863}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

@@

>>>>

\begin{abstract}
% Context:
Chemical tagging promises to use detailed abundance measurements to
identify spatially separated stars that were in fact born together
(in the same molecular cloud), long ago.
This idea has not yielded much practical success, presumably because of the
noise and incompleteness in chemical-abundance measurements.
We have succeeded in substantially improving spectroscopic
measurements with \thecannon, which has now delivered $15$ individual
abundances for $\sim 10^5$ stars observed as part of the
\apogee\ spectroscopic survey, with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars
in the 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known phase-space clusters and find other interesting structures.
This is the first-ever project to identify phase-space structures
at survey-scale by blind search purely in abundance space;
it verifies the precision of the abundance measurements delivered by \thecannon;
the prospects for future data sets appear very good.
\end{abstract}

@@

In what follows, we are going to use \apogee\ \acronym{DR12}
data \citep{apogeedr12, dr12}
from \sdssiii\ \citep{sdssiii}, in
which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---is capable of exploring all parts of the disk,
including the thin parts.
However, it has the disadvantage that its spatial coverage is incomplete
(that is, limited pointings produce a fractured sky map),
which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage \citep{galah}; that
said, it will not observe much of the thin disk.

@@

Another issue with k-means is that it effectively uses metric
distances in the $D$-space; it presumes Euclidean isotropy.
We choose here to work in the hydrogen-normalized abundance space, the
space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
But in addition to this, we re-scale these by approximate typical measurement
precisions obtained by \thecannon\ before running k-means.
These scalings were
  [C/H]/0.041,
  [N/H]/0.044,
  [O/H]/0.037,
  [Na/H]/0.111,
  [Mg/H]/0.032,
  [Al/H]/0.055,
  [Si/H]/0.041,
  [S/H]/0.054,
  [K/H]/0.069,
  [Ca/H]/0.043,
  [Ti/H]/0.072,
  [V/H]/0.146,
  [Mn/H]/0.041,
  [Fe/H]/0.019, and
  [Ni/H]/0.034.
This scaling makes the space close to isotropic in measurement uncertainty or
observational precision.
Finally and related, because it looks for clusters compact in
metric distance, k-means is more sensitive to clusters that are spherical
in the scaled abundance space than clusters of the same density
that are elongated in any sense.
Importantly we use \emph{only} abundance-space information, and no
positional or velocity information (nor $\teff$ nor $\logg$ nor any
targeting or observational meta-data) as input to the clustering
algorithm.

@@

We find a mean metallicity of [Fe/H]$ = -1.3$ for the cluster we
associate as M~5 (\figurename~\ref{fig:M5}), which agrees well with
the [Fe/H]$ = -1.33 \pm 0.03$ measurement 
\citep{Koch_2010}, and the $-1.29 \pm 0.02$ value listed in the
standard compilation (\citealt{Harris_1996}; accessed 2016).  The correlations
in light elements---specifically Mg--Al and C--N---also agree well
with other studies \citep{Ivans_2001,meszaros}.  In particular we find
only a weak correlation in the [Na/Fe]--[O/Fe] abundance ratios
\citep{Lai_2011}.  However, when we consider the extent of the
literature on M~5, it would suggest that we do not recover the full
extent of these correlations: the stars with the highest [Na/Fe],
[Mg/Al], and lowest [O/Fe] abundance ratios are not represented in the
cluster that we associate as M~5.  This is likely a consequence of our
(poor) choice of the k-means algorithm, which is most effective for
near-circular distributions in the scaled abundance space, and reduced in power for clusters
 that have polynomial relationships in dimensional
space (for example, [Na/Fe]--[O/Fe]).  Indeed, 
abundances for 122 members of M~5 have been reported in \apogee\ \acronym{DR12} data (\citealt{meszaros}),
whereas the cluster we associate as M~5 only contains $\approx60$
members, and only covers about half of the extent of [Na/Fe]--[O/Fe]
relationship.

@@

It is conceivable that this structure, being near the outskirts of the
abundance distribution, is somehow concentrated in abundance space by
some kind of shrinkage (in the statistical sense) induced by
\thecannon's regression.
However, it is a very dense feature in abundance space, and is worth
following up.
If follow-up observations show that the stars have abundances and that
are consistent with having origin in a single stellar population or
low-metallicity dwarf galaxy, it could represent an accretion event in
the Milky Way halo.

The k-means algorithm employed here is by no means optimal, and other clusters
we identify are accompanied by a few stars that are not currently co-spatial.
Those stars may be unassociated interlopers that have been misclassified,
or they may be true cluster members that are now unbound. However in the
case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
the situation is far more extreme. We identify a group of stars with
very similar abundances in 15 dimensions that are now spread throughout
the Galaxy halo.
Because some stars ($<8$\,percent) are candidate members of the massive globular
cluster M53, it provides the tantalizing possibility that these stars may
indeed be all members of the same co-natal gas cloud, implying that 
these stars have been accreted onto the halo from a single globular
cluster early in the Milky Way's formation (as
has been hypothesized elsewhere; for example in \citealt{grebel}). While speculative,
this idea demonstrates the promise of chemical tagging.

@@

\acknowledgements
It is a pleasure to thank an anonymous referee for useful comments that have led to improvements in the manuscript.
We also thank
  Joss Bland-Hawthorn (Sydney),
  Jo Bovy (Toronto),
  Charlie Conroy (Harvard),
  Katia Cunha (\acronym{NOAO}),
  Amina Helmi (Kapteyn),
  Jeremy Magland (\acronym{SCDA}),
  Don Schneider (\acronym{PSU}),
  Keivan Stassun (Vanderbilt),
  Angus Williams (Cambridge),
  and the Blanton--Hogg group meeting
for valuable discussions and comments.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1312863}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU},
  the Australian Research Council (DECRA Fellowship DE140100598),
  and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

@@

\bibitem[Martell \& Grebel(2010)]{grebel}
Martell, S.~L., \& Grebel, E.~K., 2010, \aap, 519, A14 
